Spring haul-out behavior of seals in the Bering and Chukchi seas

### 2024-02-17 analysis/paper/London_HauloutBeahvior.Rmd
lori q comments addressed

  ringed seals (_Phoca hispida_) can be out of the water but hidden from view in

@@

The global climate disruption is causing considerable reduction in Arctic sea
ice extent, volume, and seasonal presence [@kwok2018;
@meier2014; @overland2021; @wang2018]. These changes have tangible effects on
Arctic organisms, ecosystems, and the human communities who live in the region
[@huntington2020]. Such disruptions are a particular cause of concern for the
ice-associated seals that depend on spring and early summer sea ice (March-June)
in the Bering and Chukchi seas as a platform for important life history
functions, such as pupping, nursing, breeding behavior, and molting
[@boveng2009; @boveng2013a; @cameron2010; @kelly2010]. Limited data and large
knowledge gaps complicate predictions about the ultimate effects of changes in
sea ice on the behavior, health, abundance, and distribution of these seals. To
date, indices of seal health sampled during periods of declining sea ice differ
regionally [@crawford2015; @harwood2020]. Knowledge about evolutionary
constraints on the timing of reproductive and molting behavior is generally
lacking, so it is difficult to predict how or if ice-associated seal species
adapt to future changes (e.g., by adjusting pupping or molting schedules to
earlier dates or different locales). This is further complicated by the
spatio-temporal variation in the phenology of these life history events within
regions and throughout their full ranges. Additionally, trends in abundance of
these species are unknown, so it is difficult to assess the effect, if any,
declines in sea ice habitat have had, or will have, on seal demography.

@@

How ice-associated seals use sea ice as a haul-out platform varies between
species. Ribbon seals (_Histriophoca fasciata_) haul out of the water almost
exclusively on sea ice and are mostly pelagic outside the spring pupping,
breeding, and molting season [@boveng2018a]. Although spotted (_Phoca largha_)
and bearded (_Erignathus barbatus_) seals occasionally rest on coastal land,
they primarily use sea ice as a haul-out platform during the spring and early
summer [@frost2018]. Ringed seals (_Phoca hispida_) --- not included in this
study --- haul out on sea-ice but primarily use snow lairs on sea-ice during
winter and spring.

@@

For this study we used haul-out behavior data and location estimates from
bio-loggers deployed on bearded, ribbon, and spotted seals in the Bering,
Chukchi, and western Beaufort seas by multiple organizations as part of
collaborative investigations from 2005 through 2021. Seals were captured using
nets and bio-loggers were attached during studies based in coastal communities
or on research ships. Ship-based capture events occurred during spring near the
southern ice edge in the Bering Sea between 2005 and 2018. Land-based capture
events occurred between 2005 and 2020 from May to October, generally between the
coastal communities of Scammon Bay, Alaska in the Bering Sea, Utqiaġvik, Alaska
in the Chukchi Sea, and Nuiqsut in the Beaufort Sea (Supplemental
Material, [S1](#s1)). Data from additional
deployments along the Kamchatka peninsula in the western Bering Sea are also
included. We refer readers to Figure \@ref(fig:deployplot) and the primary literature
for detailed capture and bio-logger attachment methods (see also Supplemental
Material, [S1](#s1)).

@@

(ref:deployTblCaption) **Summary of bio-logger data across seal species and age classification from 1 March to 15 July 2005-2021.** \linebreak Total seal hours represents the sum of hourly observations across all seals used in the analysis. Because young-of-the-year are advanced to subadult on 1 March of the following year, some individual seals are represented in both columns in this table

@@

(ref:examplePlotCaption) **Example percent-dry actogram from bio-logger data**
\linebreak Haul-out behavior observations recorded by a bio-logger deployed on a
ribbon seal over two years during the months of March, April, and May. This
actogram plot represents the transmitted hourly percent-dry values. Not all data
during a deployment are reliably transmitted from the bio-logger and since data
are transmitted in 24-hour chunks occasional missing days are present in the
record. Here, missing data that were not successfully received from the device
are represented as empty rectangles.

@@

In addition to sex and age class, we analyzed variables that might help explain
variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and local solar hour (for diurnal effects). We calculated local solar
hour using the {solaR} package [@perpinan2012] within the R statistical
environment [@rcoreteam2021] based on the weighted daily average locations. We
also linked the weighted average daily locations to weather values from the
North American Regional Reanalysis (NARR) model produced by the National Centers
for Environmental Prediction [@mesinger2006a]. The NARR model assimilates
observational data to produce a long-term picture of weather over North America
and portions of the surrounding seas. Weather variables are made available
across the region 8 times daily. For this study, NARR weather values were limited
to the extent of our study area over the Bering, Chukchi, and Beaufort seas at 3-hr
intervals based on the native grid resolution of 32 km (1024 km^2^). The
following weather variables are known to affect haul-out behavior in
other Arctic pinnipeds [@reder2003; @udevitz2009; @perry2017] and were
interpolated and assigned to daily seal locations using a bilinear method: 1)
air temperature at 2 m above the Earth's surface, 2) wind consisting of
northerly and easterly vector components converted to wind speed using the
Euclidean norm, 3) barometric pressure at sea level, and 4) precipitation 
(Table \@ref(tab:tblCovariates)).

For all seal species, we considered the following variables when modeling the
hourly haul-out behavior: day-of-year, solar hour, temperature, wind speed,
barometric pressure, precipitation, and wind chill (represented by a
*wind:temperature* interaction [@udevitz2009]). Ribbon and spotted seal models
included age-sex class and interactions between day-of-year and age-sex class,
but we omitted these from bearded seal models due to poor representation of
age-sex classes (Table \@ref(tab:tblDeploy)). Bearded seal models included a
latitudinal effect (and an interaction with day-of-year), since bearded seals
occupy a substantial range during the spring and we were interested in possible
differences in the timing of haul-out behavior along a latitudinal gradient. We
omitted the latitudinal effect from ribbon and spotted seal models because,
during the spring, these species are most prevalent near the southern ice edge
in the Bering Sea [@conn2014a].

@@

To properly account for temporal dependence and to take advantage of
computational efficiency, we used generalized linear mixed pseudo-models
(GLMPMs; [@verhoef2010a]) to model variation in haul-out behavior as a function
of (1) covariate predictors, (2) temporally autocorrelated random effects, and
(3) individual random effects representing heterogeneity in individual behavior.
We used the glmmLDTS package [@verhoef2010a] to implement GLMPMs. We explored
two different model formulations for our data and we fit separate models to
bearded, ribbon, and spotted seal data sets as we expected differing behavior by
species. Separate models for each species were also needed because a single,
very large data set proved computationally intractable. In our first model
formulation and for each species, we fitted a year-independent model that
predicted average haul-out behavior as a function of demographic, weather,
seasonal, and diurnal effects. Second, for ribbon and spotted seals (which had
considerably more data than bearded seals), we fitted models that included all
the effects from the first model, but also permitted annual variation in
haul-out timing. This second set of models was used to examine whether
haul-out patterns varied by year and to determine the annual timing of apparent
peaks in haul-out behavior. For both models, we assumed an hourly Bernoulli
response (a binary 0-1 response dependent upon whether the tag was greater than 
50% dry for a given hour) where the linear predictor was 
modeled on the logit scale. This is
consistent with previous approaches [@verhoef2014a, @london2012] and only
`r (1 - (portion_10(analysis_data) + portion_90(analysis_data))) * 100`% of our
observations fell between 10% and 90% hourly percent-dry.

@@

We recognize that the species and ecosystems we studied are within the ancestral
and present-day environs of the Inpuiat and Yup'ik people who, through many
uncredited contributions of traditional knowledge, provided early western
naturalists and scientists with much of what gets described as the ‘basic
biology’ of Arctic seals. The deployment of bio-logging devices used in this
study were often done in collaboration with Alaska Native seal hunters near
their communities. We would like to especially acknowledge the communities of
Kotzebue, Koyuk, Nome, Nuiqsut, Scammon Bay, St. Michael, Utqiaġvik, and Ulguniq
(Wainwright) and the following individuals: James Adams, Jeff Barger, David
Barr, Wendell Booth, Cyrus Harris, Nereus 'Doc' Harris, Grover Harris, Lee
Harris, Tom Jones, Frank Garfield, Brenda Goodwin, Henry Goodwin, John Goodwin,
Pearl Goodwin, Willie Goodwin, Brett Kirk, Noah Naylor, Virgil Naylor Jr.,
Virgil Naylor Sr., Dan Savetilik, Chuck Schaeffer, Ross Schaeffer, Allen Stone,
and Randy Toshavik from Kotzebue, Alaska; Merlin Henry from Koyuk, Alaska; Tom
Gray from Nome; Vernon Long and Richard Tukle from Nuiqsuit, Alaska; Morgan
Simon, River Simon, and Al Smith from Scammon Bay, Alaska; Alex Niksik Jr. from
St. Michael, Alaska; Billy Adams, James Aiken, Tim Aiken, Howard Kittick,
Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, and Joe Skin from Utqiaġvik,
Alaska; Mary Ellen Ahmaogak, Enoch Oktollik, Shawn Oktollik, Stacey Osborn, and
Fred Rexford from Ulguniq, Alaska.

@@

Telemetry data from the Alaska Department of Fish and Game (ADFG) and the North
Slope Borough Department of Wildlife Management (NSB) were important
contributions to the findings presented here. Deployments in the western Bering
Sea were done in collaboration with Russian colleagues and North Pacific
Wildlife.

The findings and conclusions in the paper are those of the author(s) and do not
necessarily represent the views of the National Marine Fisheries Service, NOAA.
Any use of trade, product, or firm names does not imply an endorsement by the
U.S. Government. Funding for this study was provided by the U.S. National
Oceanic and Atmospheric Administration. The field work was conducted under the
authority of Marine Mammal Protection Act Research Permits Nos. 782-1676,
782-1765, 15126, and 19309 issued by the National Marine Fisheries Service, and
Letters of Assurance of Compliance with Animal Welfare Act regulations, Nos.
A/NW 2010-3 and A/NW 2016-1 from the Alaska Fisheries Science Center/Northwest
Fisheries Science Center Institutional Animal Care and Use Committee (IACUC). ).
Funding to ADFG for tagging seals was provided by the Bureau of Ocean Energy
Management (No. M13PC0015 for work in Kotzebue, Alaska in 2009) and the Office
of Naval Research (No. N00014-16-1-3019). ADFG and NSB field work was covered by
Research Permits Nos. 358-1585, 358-1787, 15324, and 20466 and by ADF&G IACUC
permits Nos. 06-16, 09-21, 2014-03, 2015-25, 2016-23, 0027-2017-27,
0027-2018-29, 0027-2019-041.

@@

The following series of figures (\@ref(fig:beardedPredSE), \@ref(fig:ribbonPredSE),
and \@ref(fig:spottedPredSE)) show the seasonal variability in predicted haul-out
probability and the associated 95% confidence intervals for bearded, ribbon, and
spotted seals. The predictions shown are based on the same data used in
\@ref(fig:beardedHOCal), \@ref(fig:ribbonHOCal), and \@ref(fig:spottedHOCal) but
selected for three to local solar hours (07:00, 12:00, and 17:00) so the 
confidence intervals can also be shown and comparisons can be made.

@@

During the peer review process for this manuscript, Anthony Fischbach suggested
the possibility of using predicted insolation (or solar radiation) values from
the reanalysis model as a more direct and, potentially, more informative
predictor of the daily haul-out cycle in seals compared to time of day. The
notion being that seals are, likely, directly responding to changes in solar radiation
throughout the day and not what time of day it is (i.e. seals don't have
watches). Additionally, given the energetic benefits of increased solar
radiation it could be more informative as we would expect seals might have a
higher haul-out probability on sunnier days and for there to be geographic
variability in haul-out behavior associated with geographical differences in
insolation. This approach has an additional benefit of being more parsimonious
compared to our use of the Fourier series or other approaches to represent
hour-of-day in the model (e.g. 24 factors for each hour).

>>>>

  ringed seals (_Phoca (pusa) hispida_) can be out of the water but hidden from view in

@@

Global climate disruption is causing considerable reduction in Arctic sea
ice extent, volume, and seasonal presence [@kwok2018;
@meier2014; @overland2021; @wang2018]. These changes have tangible effects on
Arctic organisms, ecosystems, and the people who live in the region
[@huntington2020]. Such disruptions are a particular cause of concern for the
ice-associated seals that depend on spring and early summer sea ice (March-June)
in the Bering and Chukchi seas as a platform for important life history
functions, such as pupping, nursing, breeding behavior, and molting
[@boveng2009; @boveng2013a; @cameron2010; @kelly2010]. Limited data and large
knowledge gaps complicate predictions about the ultimate effects of changes in
sea ice on the behavior, health, abundance, and distribution of these seals. To
date, indices of seal health sampled during periods of declining sea ice differ
regionally [@crawford2015; @harwood2020]. Knowledge about evolutionary
constraints on the timing of reproductive and molting behavior is generally
lacking, so it is difficult to predict how or if ice-associated seal species
will adjust to future changes (e.g., by adjusting pupping or molting schedules to
earlier dates or different locales). This is further complicated by the
spatio-temporal variation in the phenology of these life history events within
regions and throughout their full ranges. Additionally, trends in abundance of
these species are unknown, so it is difficult to assess the effect, if any,
declines in sea ice habitat have had, or will have, on seal demography.

@@

How ice-associated seals use sea ice as a haul-out platform varies among
species. Ribbon seals (_Histriophoca fasciata_) haul out of the water almost
exclusively on sea ice and are mostly pelagic outside the spring pupping,
breeding, and molting season [@boveng2018a]. Although spotted (_Phoca largha_)
and bearded (_Erignathus barbatus_) seals occasionally rest on coastal land,
they primarily use sea ice as a haul-out platform during the spring and early
summer [@frost2018]. Ringed seals (_Phoca (pusa) hispida_) --- not included in this
study --- haul out on sea-ice but primarily use snow lairs on sea-ice during
winter and spring.

@@

For this study we used haul-out behavior data and location estimates from
bio-loggers deployed on bearded, ribbon, and spotted seals in the Bering,
Chukchi, and western Beaufort seas by multiple organizations as part of
collaborative investigations from 2005 through 2021. Seals were captured using
nets and bio-loggers were attached during studies based in coastal communities
or on research ships. Ship-based capture events occurred during spring near the
southern ice edge in the Bering Sea between 2005 and 2018. Land-based capture
events occurred between 2005 and 2020 from May to October, generally between the
Alaska coastal communities of Scammon Bay in the Bering Sea, Utqiaġvik
in the Chukchi Sea, and Nuiqsut in the Beaufort Sea (Supplemental
Material, [S1](#s1)). Data from additional
deployments along the Kamchatka peninsula in the western Bering Sea are also
included. We refer readers to Figure \@ref(fig:deployplot) and the primary literature
for detailed capture and bio-logger attachment methods (see also Supplemental
Material, [S1](#s1)).

@@

(ref:deployTblCaption) **Summary of bio-logger data by seal species and age classification from 1 March to 15 July 2005-2021.** \linebreak Seal hours represents the sum of hourly observations across all seals used in the analysis. Because young-of-the-year are advanced to subadult on 1 March of the following year, some individual seals are represented in both columns in this table

@@

(ref:examplePlotCaption) **Example percent-dry actogram from bio-logger data**
\linebreak Haul-out behavior observations recorded by a bio-logger deployed on a
ribbon seal over two years during the months of March, April, and May. This
actogram plot represents the transmitted hourly percent-dry values. Not all data
during a deployment are reliably transmitted from the bio-logger and because data
are transmitted in 24-hour chunks occasional missing days are present in the
record. Here, missing data that were not successfully received from the device
are represented as empty rectangles.

@@

In addition to sex and age class, we analyzed variables that might help explain
variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and local solar hour (for diurnal effects). We calculated local solar
hour using the {solaR} package [@perpinan2012] within the R statistical
environment [@rcoreteam2021] based on the weighted daily average locations. We
also linked the weighted average daily locations to weather values from the
North American Regional Reanalysis (NARR) model produced by the National Centers
for Environmental Prediction [@mesinger2006a]. The NARR model assimilates
observational data to produce a long-term picture of weather over North America
and portions of the surrounding seas. Weather variables are made available
across the region 8 times daily. For this study, NARR weather values were limited
to the extent of our study area over the Bering, Chukchi, and Beaufort seas at 3-hr
intervals based on the original grid resolution of 32 km (1024 km^2^). The
following weather variables are known to affect haul-out behavior in
other Arctic pinnipeds [@reder2003; @udevitz2009; @perry2017] and were
interpolated and assigned to daily seal locations using a bilinear method: 1)
air temperature at 2 m above the Earth's surface, 2) wind consisting of
northerly and easterly vector components converted to wind speed using the
Euclidean norm, 3) barometric pressure at sea level, and 4) precipitation 
(Table \@ref(tab:tblCovariates)).

For all seal species, we considered the following variables when modeling the
hourly haul-out behavior: day-of-year, solar hour, temperature, wind speed,
barometric pressure, precipitation, and wind chill (represented by a
*wind:temperature* interaction [@udevitz2009]). Ribbon and spotted seal models
included age-sex class and interactions between day-of-year and age-sex class,
but we omitted these from bearded seal models due to poor representation of
age-sex classes (Table \@ref(tab:tblDeploy)). Bearded seal models included a
latitudinal effect (and an interaction with day-of-year), because bearded seals
occupy a substantial range during the spring and we were interested in possible
differences in the timing of haul-out behavior along a latitudinal gradient. We
omitted the latitudinal effect from ribbon and spotted seal models because,
during the spring, these species are most prevalent near the southern ice edge
in the Bering Sea [@conn2014a].

@@

To properly account for temporal dependence and to take advantage of
computational efficiency, we used generalized linear mixed pseudo-models
(GLMPMs; [@verhoef2010a]) to model variation in haul-out behavior as a function
of (1) covariate predictors, (2) temporally autocorrelated random effects, and
(3) individual random effects representing heterogeneity in individual behavior.
We used the glmmLDTS package [@verhoef2010a] to implement GLMPMs. We explored
two different model formulations for our data and we fit separate models to
bearded, ribbon, and spotted seal data sets as we expected differing behavior by
species. Separate models for each species were also needed because a single,
very large data set proved computationally intractable. In our first model
formulation and for each species, we fitted a year-independent model that
predicted average haul-out behavior as a function of demographic, weather,
seasonal, and diurnal effects. Second, for ribbon and spotted seals (which had
considerably more data than bearded seals), we fitted models that included all
the effects from the first model, but also permitted annual variation in
haul-out timing. This second set of models was used to examine whether
haul-out patterns varied by year and to determine the annual timing of apparent
peaks in haul-out behavior. For both models, we assumed an hourly Bernoulli
response (a binary 0-1 response dependent upon whether the tag was greater than 
50% dry for a given hour) where the linear predictor was 
modeled on the logit scale. This is
consistent with previous approaches [@verhoef2014a; @london2012] and only
`r (1 - (portion_10(analysis_data) + portion_90(analysis_data))) * 100`% of our
observations fell between 10% and 90% hourly percent-dry.

@@

We recognize that the species and ecosystems we studied are within the ancestral
and present-day environs of the Inpuiat and Yup'ik people who, through many
uncredited contributions of traditional knowledge, provided early western
naturalists and scientists with much of what gets described as the ‘basic
biology’ of Arctic seals. The deployment of bio-logging devices used in this
study were often done in collaboration with Alaska Native seal hunters and the
approval of
their communities. We would like to especially acknowledge the communities of
Kotzebue, Koyuk, Nome, Nuiqsut, Scammon Bay, St. Michael, Utqiaġvik, and Ulguniq
(Wainwright) and the following individuals: James Adams, Jeff Barger, David
Barr, Wendell Booth, Cyrus Harris, Nereus 'Doc' Harris, Grover Harris, Lee
Harris, Tom Jones, Frank Garfield, Brenda Goodwin, Henry Goodwin, John Goodwin,
Pearl Goodwin, Willie Goodwin, Brett Kirk, Noah Naylor, Virgil Naylor Jr.,
Virgil Naylor Sr., Dan Savetilik, Chuck Schaeffer, Ross Schaeffer, Allen Stone,
and Randy Toshavik from Kotzebue; Merlin Henry from Koyuk; Tom
Gray from Nome; Vernon Long and Richard Tukle from Nuiqsuit; Morgan
Simon, River Simon, and Al Smith from Scammon Bay; Alex Niksik Jr. from
St. Michael; Billy Adams, James Aiken, Tim Aiken, Howard Kittick,
Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, and Joe Skin from Utqiaġvik,
Alaska; Mary Ellen Ahmaogak, Enoch Oktollik, Shawn Oktollik, Stacey Osborn, and
Fred Rexford from Ulguniq.

@@

Telemetry data from the Alaska Department of Fish and Game (ADF&G) and the North
Slope Borough Department of Wildlife Management (NSB) were important
contributions to the findings presented here. Deployments in the western Bering
Sea were done in collaboration with Russian colleagues and North Pacific
Wildlife.

The findings and conclusions in the paper are those of the author(s) and do not
necessarily represent the views of the National Marine Fisheries Service, NOAA.
Any use of trade, product, or firm names does not imply an endorsement by the
U.S. Government. Funding for this study was provided by the U.S. National
Oceanic and Atmospheric Administration. The field work was conducted under the
authority of Marine Mammal Protection Act Research Permits Nos. 782-1676,
782-1765, 15126, and 19309 issued by the National Marine Fisheries Service, and
Letters of Assurance of Compliance with Animal Welfare Act regulations, Nos.
A/NW 2010-3 and A/NW 2016-1 from the Alaska Fisheries Science Center/Northwest
Fisheries Science Center Institutional Animal Care and Use Committee (IACUC). ).
Funding to ADF&G for tagging seals was provided by the Bureau of Ocean Energy
Management (No. M13PC0015 for work in Kotzebue, Alaska in 2009) and the Office
of Naval Research (No. N00014-16-1-3019). ADF&G and NSB field work was covered by
Research Permits Nos. 358-1585, 358-1787, 15324, and 20466 and by ADF&G IACUC
permits Nos. 06-16, 09-21, 2014-03, 2015-25, 2016-23, 0027-2017-27,
0027-2018-29, 0027-2019-041.

@@

The following series of figures (\@ref(fig:beardedPredSE), \@ref(fig:ribbonPredSE),
and \@ref(fig:spottedPredSE)) show the seasonal variability in predicted haul-out
probability and the associated 95% confidence intervals for bearded, ribbon, and
spotted seals. The predictions shown are based on the same data used in
\@ref(fig:beardedHOCal), \@ref(fig:ribbonHOCal), and \@ref(fig:spottedHOCal) but
selected for three local solar hours (07:00, 12:00, and 17:00) so the 
confidence intervals can also be shown and comparisons can be made.

@@

During the peer review process for this manuscript, Anthony Fischbach suggested
the possibility of using predicted insolation (or solar radiation) values from
the reanalysis model as a more direct and, potentially, more informative
predictor of the daily haul-out cycle in seals compared to time of day. The
notion being that seals are, likely, directly responding to changes in solar radiation
throughout the day and not what time of day it is (i.e. seals don't have human
watches). Additionally, given the energetic benefits of increased solar
radiation it could be more informative as we would expect seals might have a
higher haul-out probability on sunnier days and for there to be geographic
variability in haul-out behavior associated with geographical differences in
insolation. This approach has an additional benefit of being more parsimonious
compared to our use of the Fourier series or other approaches to represent
hour-of-day in the model (e.g. 24 factors for each hour).
